---
title: "Basic Auto-scaling<br><small>Évaluation de Performances 2024, TD #2, INFO4, Polytech Grenoble</small>"
author: "Jonatha Anselmi"
date: "2024-02-22"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Auto-scaling

Auto-scaling is a powerful mechanism for the efficient management of serverless, edge and cloud computing systems, both from the perspective of user-perceived performance and overall power consumption. It allows the *net* service capacity to scale up or down in response to the current load and the latter distributes incoming jobs across the set of available servers. 

In auto-scaling systems, Servers can be in three states: warm, cold and initializing. A server is said \emph{warm} if turned on, \emph{cold} if turned off and \emph{initializing}
if making the transition from cold to warm.
An initializing server performs basic startup operations such as connecting to database, loading libraries, etc. This is the time to provision a new function instance.
Only warm servers are allowed to receive jobs.
A server is also said \emph{idle-on} if warm but not processing any job, and \emph{busy} if warm and processing some job.
Typically, billing policies charge per number of warm and initializing servers used per time unit.

Our aim is to simulate the dynamics of some basic auto-scaling models.


### M/M/K/K queueing model

A first basic model for autoscaling is given by a $M/M/K/K$ queue. Here, the interpretation is that $N$ represents the nominal capacity, i.e., the maximum number of servers that can be up and running. Then, if $n_t$ is the number of jobs in the system at time $t$, then $n_t$ also represents the number of busy servers. Therefore, $K-n_t$ may model the number of cold servers, provided that

  1. a server becomes cold immediately after processing a job (or equivalently, the expiration rate is very high so that it can be ignored).
  
  2. servers cannot enter the initialization phase (or equivalently, the initialization rate is very high so that it can be ignored).

The $M/M/K/K$ queue can be described by a continuous-time Markov chain with state space $S=\{0,1,\ldots,N\}$ and transition matrix $Q=(q_{i,j})$ where all the non-zero elements are $q_{i,i+1}=\lambda$ for all $i=0,\ldots,K-1$ and $q_{i,i-1}=i \mu $ for all $i=1,\ldots,K$.

Let us write a code to simulate the M/M/N/N queue.

```{r eval=TRUE, include=TRUE}
knitr::opts_chunk$set(echo = TRUE)

simulate_MMNN <- function(lambda, mu, K){
  
  set.seed(17); # Set seed for reproducibility
  N=1e5; # number of events to simulate;

  p_blocking=0; # blocking probability
  p_blocking_transient=rep(N, 0); # transient blocking probability
  T=0; # simulation time
  
  state=0;
  
  for (i in 1:N) {
  
    U=lambda+state*mu;
    time_in_state=rexp(1,U)
    T=T+time_in_state;
  
    if (state==K) {
      # For the moment, let us find the overall time spent in state K
      p_blocking=p_blocking+time_in_state;
    }
  
    p_blocking_transient[i]=p_blocking/T;
  
    if (runif(1)<lambda/U){
        # An arrival occurred
        if (state<K){
          # The incoming job has room so it is accepted
          state=state+1;
        }
    } else {
        # A departure occurred
        state=state-1;    
    }
  }
  
  # The blocking probability by simulation
  p_blocking=p_blocking/T;
  
  # The blocking probability by theory (The Erlang B formula)
  p_blocking_theory=0; 
  normalizing_constant=0;
  for (i in 0:K){
    normalizing_constant = normalizing_constant + (lambda/mu)**i / factorial(i);
  }
  
  p_blocking_theory = (lambda/mu)**K / factorial(K);
  p_blocking_theory = p_blocking_theory / normalizing_constant;
  
  
  # Print the blocking probabilities (simulation and theory)
  events=1:N;
  plot(events, p_blocking_transient, type = "l", lty = 1)
  abline(h = p_blocking_theory, col="red")
  print(p_blocking)
  print(p_blocking_theory)
}
```


Let's simulate.

```{r eval=TRUE, include=TRUE}
knitr::opts_chunk$set(echo = FALSE)

set.seed(11); # Set seed for reproducibility
K=10;         # number of servers
lambda=0.6*K; # arrival rate
mu=1;         # service rate

simulate_MMNN(lambda,mu,K);

```


### M/M/K/K queueing model + expiration rate

Let us now go further. Building on the M/M/K/K queue, we want to add the feature that when a server becomes idle-on, it becomes cold by itself after a scale down delay, or \emph{expiration time}, provided that during such time the server received no job. This scale-down rule is common in serverless computing platforms.

To model this, 

The $M/M/K/K$ queue can be described by a continuous-time Markov chain with state space $S=\{0,1,\ldots,N\}$ and transition matrix $Q=(q_{i,j})$ where all the non-zero elements are $q_{i,i+1}=\lambda$ for all $i=0,\ldots,K-1$ and $q_{i,i-1}=i \mu $ for all $i=1,\ldots,K$.

Let us write a code to simulate this Markov chain in continuous time.

<br>