---
title: "Practicing with continuous-time Markov chain and queues<br><small>Évaluation de Performances 2025, TD #6, INFO4, Polytech Grenoble</small>"
author: "Jonatha Anselmi"
date: "2025-02-28"
output: html_document
---

### Problem statement

The aim of this session is to practice working with continuous-time Markov chains in R. To achieve this, we write code to simulate the dynamics of a basic queueing system.  


#### Assumptions

Arrival to the queueing system follow a Poisson process with rate $\lambda$, and
job service times are random variables that follow an exponential distribution with rate $\mu$.


### M/M/1

We start by simulating the number of jobs in a M/M/1 queue. To do:

1. Draw the transition diagram on a piece of paper.

2. Plot a sample trajectory of the number of jobs

3. Print the average number of jobs in the systems, say $Q_{sim}$.

4. Compare $Q_{sim}$ with its theoretical formula.


```{r eval=TRUE, include=TRUE}
knitr::opts_chunk$set(echo = TRUE)

# This function performs the required simulation.

N=1e5; # overall number of events (departure or arrivals)

lambda=0.7; mu=1; # arrival and service rates

state=0; # set the initial state
Qsim=0;  # average number of jobs in the system

trajectory=c();
events=c();
for (i in 1:N) {

    # which events can occur? when?
    time_2_next_arr=rexp(1,lambda);
    time_2_next_dep=rexp(1,mu);
    trajectory=c(trajectory, state);
    if(state>0) {
        
        Qsim=Qsim+min(time_2_next_dep,time_2_next_arr)*state;
        events=c(events, min(time_2_next_dep,time_2_next_arr));
        
        if (time_2_next_dep<time_2_next_arr) {
            # a departure occurred
             state=state-1;
        } else {
             # an arrival occurred
             state=state+1;
        }
    } else {
        Qsim=Qsim+time_2_next_arr*state; # useless because state==0
        events=c(events, time_2_next_arr);
        state=state+1;    
    }
    
} #for

rho=lambda/mu;
cat("Load rho = ",rho,"\n");
cat("Average number of jobs (simulation) =  ", Qsim/sum(events),'\n');
cat("Average number of jobs (theory)     =  ", rho/(1-rho),'\n');

plot(cumsum(events),trajectory, type = 'S', xlab="time", ylab="number of jobs")

```


### M/M/1 with batches

We consider again the M/M/1 queue but this time we assume that jobs can arrive in pairs, i.e., one job arrives with probability $p$ and two jobs arrive with probability $1-p$ -- note that this is not an M/M/1 queue.
Again:

1. Draw the transition diagram on a piece of paper.

2. Plot a sample trajectory of the number of jobs

3. Print the average number of jobs in the systems, say $Q_{sim}$.

How does $Q_sim$ compare with the mean number of jobs of an M/M/1 queue with the same arrival and service rates?

```{r eval=TRUE, include=TRUE}
knitr::opts_chunk$set(echo = TRUE)

# This function performs the required simulation.

N=1e3; # overall number of events (departure or arrivals)

lambda=0.7; mu=1; # arrival and service rates
p=0.7;
state=0; # set the initial state
Qsim=0;  # average number of jobs in the system

trajectory=c();
events=c();
for (i in 1:N) {

    # which events can occur? when?
    time_2_next_arr1=rexp(1,lambda*p);
    time_2_next_arr2=rexp(1,lambda*(1-p));
    time_2_next_dep=rexp(1,mu);
    trajectory=c(trajectory, state);
    if(state>0) {
        
        Qsim=Qsim+min(time_2_next_dep,time_2_next_arr1,time_2_next_arr2)*state;
        events=c(events, min(time_2_next_dep,time_2_next_arr1,time_2_next_arr2));
        
        if (time_2_next_dep<min(time_2_next_arr1,time_2_next_arr2)) {
            # a departure occurred
             state=state-1;
        }
        if (time_2_next_arr1<min(time_2_next_dep,time_2_next_arr2)) {
             # one arrival occurred
             state=state+1;
        }
        if (time_2_next_arr2<min(time_2_next_dep,time_2_next_arr1)) {
             # two arrivals occurred
             state=state+2;
        }
    } else {
        events=c(events, min(time_2_next_arr1,time_2_next_arr2));
        
        if (time_2_next_arr1<time_2_next_arr2) {
             # one arrival occurred
             state=state+1;
        }
        if (time_2_next_arr2<time_2_next_arr1) {
             # two arrivals occurred
             state=state+2;
        }
    }
    
} #for

rho=lambda*(p+2*(1-p))/mu;
cat("Average number of jobs (simulation, rho=",rho,") =  ", Qsim/sum(events),'\n');

plot(cumsum(events),trajectory, type = 'S', xlab="time", ylab="number of jobs")
```

### Homework

Adapt the code above to allow for a more general arrival process, i.e., $k$ jobs arrive with probability $p_k$, for $k=1,\ldots,K$.

<br><br>
