---
title: "Scheduling disciplines in the G/G/1 queue with soft deadlines<br><small>Évaluation de Performances 2025, TD #2 and #3, INFO4, Polytech Grenoble</small>"
author: "Jonatha Anselmi"
date: "2025-01-24"
output: html_document
---

### Problem statement

The aim of this session is to be able to simulate and compare the dynamics of a G/G/1 queue adopting a number of scheduling disciplines and where jobs have soft deadlines. Specifically, when a job arrives, it should be completed before a deadline that is specific to the job.
The deadline is "soft" in the sense it may not be respected, though this wold imply some penalty.

We consider the following scheduling disciplines: 

1. **FCFS**: jobs are processed following the order of their arrival
    
2. **EDF** (Earliest Deadline First): jobs are processed according to the earliest deadline with preemption, i.e., the current job is interrupted if a new job with an earlier deadline  arrives.
    
3. **EDF-NP** (Earliest Deadline First Non-preemptive): As EDF but jobs are not interrupted once their processing has started.


We are interested in comparing the performance of the previous scheduling disciplines with respect to:

1. **Response Time** (RT): the mean time that a job spends in the system (in the queue and processing phases)
    
2. **Missed Deadline Ratio** (MDR): the proportion of jobs that completed after their declared deadline
    
3. **Impossible Deadline Ratio** (IDR): the proportion of jobs that declared an impossible deadline, i.e., no discipline can fulfill the deadline requirement in any scenario. This is the case of a job whose service time is larger than its declared deadline.

Towards this purpose, the objective is to create a table of the form:

|Discipline| RT  	|  MDR 	|  IDR 	|
|:---	|:---	|:---	|:---	|
|FCFS|   	|   	|   	|
|EDF|   	|   	|   	|
|EDF-NP|   	|   	|   	|

We will rely on this table to comment on the results.

#### Parameter settings

We will consider two parameter settings, referred to as Scenarios 1 and 2.
Specifically,

1. in Scenario 1, assume that service times are exponentially distributed (with rate, say, $\mu$) and deadlines are $x=5$ times larger than service times. In this case, note that IDR=0.

2. in Scenario 2, assume that service times are exponentially distributed (with rate, say, $\mu$) and deadlines are also exponentially distributed with rate $\gamma=\mu/5$. Also, deadlines are independent random variables.

In both scenarios, assume a Poisson arrival process, i.e., job interarrival times are independent and exponentially distributed with rate, say, $\lambda$.
The rate $\lambda$ is such that the system load $\rho=\lambda/\mu$ is 0.9.



#### Hints

Adapt [the code for the simulation of the G/G/1 queue](https://rpubs.com/janselmi/GG1_scheduling_disciplines).

The R function <code>which.min(v)</code> returns the index of the minimum element in vector <code>v</code>.

### Main code

Here is the main R function that performs the simulation and print the results.

```{r eval=TRUE, include=TRUE}
knitr::opts_chunk$set(echo = TRUE)

# This function performs the required simulation.
# As input, it takes the job service times (Service), their expected value (ES), the job deadlines (relative to the absolute job arrival time), and the system load
run_simulation <- function(Service, ES, Deadline, rho){
    debug=0;
    scheduling_disciplines = c("FCFS",
                               "EDF");
    num_arrival_rates=1;
    lambda = rho/ES;
    
    # Matrix of average response times for all disciplines and arrival rates
    Results <- matrix(0, nrow = length(scheduling_disciplines), ncol = 3);
    rownames(Results) <- scheduling_disciplines;
    colnames(Results) <- c("RT", "MDR", "IDR"); # performance metrics
    
    # Let's assume a Poisson process coming in, otherwise change the next line
    Arrival= cumsum(rexp(n=N, rate = lambda)); # Arrival times
    AbsDeadline = Arrival+Deadline;
    
    for (scheduling_discipline in scheduling_disciplines)
    {
        t = 0; # simulation time
        Remaining = rep(N, x=NA);    # Remaining service times of each job
        Completion = rep(N, x=NA);   # Completion time of each job
        
        CurrentTask = NA;
        NextArrival = 1;
        while (TRUE) {

if (debug==1){
    print("*********************");
    cat("Arrival:     ",Arrival,'\n');
    cat("AbsDeadline: ",AbsDeadline,'\n');
    cat("Service:     ",Service,'\n');
    cat("Remaining:   ",Remaining,'\n');
}
            dtA = NA; dtC = NA;
            if(length(Arrival[Arrival>t])>0) { # if an arrival exists after t
                dtA = head(Arrival[Arrival>t],n=1)-t ; # time to next arrival
            }
            if(!is.na(CurrentTask)) { # if a task is running
                dtC = Remaining[CurrentTask]; # time to next completion
            }
            if(is.na(dtA) & is.na(dtC)) {
                break;
            } 
        
            dt = min(dtA,dtC,na.rm=T);
        
            # update system variables
            t = t + dt;
            
            # Job arrival
            if((NextArrival <=N) & (Arrival[NextArrival] == t)) {
                Remaining[NextArrival] = Service[NextArrival];
                NextArrival = NextArrival + 1;
            }
        
            # Job departure
            if(!is.na(CurrentTask)) {
                Remaining[CurrentTask] = Remaining[CurrentTask] - dt;
                if(Remaining[CurrentTask] <= 0) {
                    # CurrentTask completed
                    Completion[CurrentTask] = t;
                    Remaining[CurrentTask] = NA;
                    CurrentTask = NA;
                }
            }
            
            # Scheduling discipline: Let's find the new task to process
            
            WaitingList=(1:NextArrival)[!is.na(Remaining)];
            
if (debug==1){
    print(paste("Sim. time:", t));
    print(paste("# of jobs arrived: ", NextArrival));
    cat("WaitingList:", WaitingList, "\n");
}
            if(length(WaitingList)>0) {
        
              if (scheduling_discipline=="FCFS"){
                CurrentTask = head(WaitingList,n=1);
              }
              
              if (scheduling_discipline=="EDF"){
                 # Solution 1
                 # CurrentTask = which(AbsDeadline==min(AbsDeadline[WaitingList]));
                 # Solution 2
                  CurrentTask = WaitingList[which.min(AbsDeadline[WaitingList])];
                 # cat(CurrentTask, " while FCFS says ", head(WaitingList,n=1), '\n');
              }
        
            }
    
if (debug==1){
    print(paste("Current Task = ", CurrentTask))
    #    readline(prompt="Press [enter] to proceed")
}
        
    } # while
    
        IDR=mean(Service>Deadline);
        MDR=mean(Completion>AbsDeadline);
        ResponseTime = mean(Completion-Arrival); 

        # Simulation completed.
        
        Results[scheduling_discipline,1]=ResponseTime;
        Results[scheduling_discipline,2]=MDR;
        Results[scheduling_discipline,3]=IDR;
        
        cat(scheduling_discipline," completed.\n");
    
    } # loop scheduling discipline
    
    cat("\nResults:\n")
    print(Results)
    
} # end function


```

### Scenario 1

In Scenario 1, service times are exponentially distributed with rate $\mu$  and deadlines are $x=5$ times larger than service times.

```{r eval=TRUE, include=TRUE}
knitr::opts_chunk$set(echo = FALSE)

set.seed(11); # Set seed for reproducibility

N = 1e4; # number of jobs to simulate;

mu=1; Service=rexp(N,rate=mu); # Service times
x=5;  Deadline=Service*x; # Relative deadlines

rho=0.9; # load
run_simulation(Service, 1/mu, Deadline, rho);

```

### Scenario 2

In Scenario 2, service times are exponentially distributed with rate $\mu$, and deadlines are independent and exponentially distributed with rate $\gamma=\mu/5$.


```{r eval=TRUE, include=TRUE}
knitr::opts_chunk$set(echo = FALSE)

set.seed(11); # Set seed for reproducibility

N = 1e4; # number of jobs to simulate;

# TO DO

```





<br><br>
