---
title: "Scheduling Disciplines in the G/G/1 queue"
author: "Jonatha Anselmi"
date: "2024-01-19"
output: html_document
---

This code simulates the behavior of a G/G/1 queue with a number of scheduling disciplines.
Then, their response time is printed.


```{r eval=TRUE, include=TRUE}
knitr::opts_chunk$set(echo = TRUE)

# This function plot the average response time of a M/G/1 queue for various scheduling disciplines.
# As input, it takes the job service times (Service) and their mean value (ES)
plot_response_times <- function(Service, ES){

debug=0;
    scheduling_disciplines = c(1:4); # 'FCFS'=1,'LCFS'=2,'ROS'=3,'SRPT'=4
    num_arrival_rates=9;
    lambdas = ES * c(1:num_arrival_rates)/(num_arrival_rates+1) ;
    
    # Matrix of average response times for all disciplines and arrival rates
    R_time <- matrix(0, nrow = length(scheduling_disciplines), ncol = length(lambdas));
    rownames(R_time) <- c("FCFS","LCFS","ROS","SRPT");
    
    cnt=0;
    for (lambda in lambdas)
    {
    cnt=cnt+1;
    Arrival= cumsum(rexp(n=N, rate = lambda)); # Arrival times
    for (scheduling_discipline in scheduling_disciplines)
    {
      t = 0; # simulation time
      Remaining = rep(N, x=NA);    # Remaining service times of each job
      Completion = rep(N, x=NA);   # Completion time of each job
      
      AvgJobs = 0;
      CurrentTask = NA;
      NextArrival = 1;
      while (TRUE) {
      
      if (debug==1){
          print("*********************");
          print(Arrival);
          print(Service);
          print(Remaining);
      }
    
          dtA = NA;
          dtC = NA;
          if(length(Arrival[Arrival>t])>0) { # if an arrival exists after t
              dtA = head(Arrival[Arrival>t],n=1)-t ; # time to next arrival
          }
          if(!is.na(CurrentTask)) { # if a task is running
              dtC = Remaining[CurrentTask]; # time to next completion
          }
          if(is.na(dtA) & is.na(dtC)) {
              break;
          } 
      
          dt = min(dtA,dtC,na.rm=T);
      
          # update system variables
          t = t + dt;
    
      if (debug==1){
          print(paste("Sim. time:", t));
          print(paste("# of jobs arrived: ", NextArrival));
      }
          # Job arrival
          
          if((NextArrival <=N) & (Arrival[NextArrival] == t)) {
              Remaining[NextArrival] = Service[NextArrival];
              NextArrival = NextArrival + 1;
          }
    
          # Job departure
          
          if(!is.na(CurrentTask)) {
              Remaining[CurrentTask] = Remaining[CurrentTask] - dt ;
              if(Remaining[CurrentTask] <= 0) {
                  # CurrentTask completed
                  Completion[CurrentTask] = t;
                  Remaining[CurrentTask] = NA;
                  CurrentTask = NA;
              }
          }
          
          # Scheduling discipline
          
          WaitingList=(1:NextArrival)[!is.na(Remaining)];
         
          if(length(WaitingList)>0) {
    
            # FCFS
            if (scheduling_discipline==1){
              CurrentTask = head(WaitingList,n=1);
            }
            
            # LCFS
            if (scheduling_discipline==2){
              # We implement the non-preemptive version of LCFS: jobs are never killed/resumed
              if(is.na(CurrentTask)) {
              # here, a task is not running, so we find a new job to serve
                CurrentTask = tail(WaitingList,n=1);
              }
            }
            
            # ROS: Random Order of Service
            if (scheduling_discipline==3){
              if(is.na(CurrentTask)) {
              # here, a task is not running, so we find a new job to serve
                if (length(WaitingList)>1) {
                  CurrentTask = sample(WaitingList,size=1);
                } else{
                  CurrentTask = WaitingList;
                }
              }
            }
            
            # SRPT: Shortest Remaining Processing Time
            if (scheduling_discipline==4){
                CurrentTask = which.min(Remaining);
            }
    
          }
    
      if (debug==1){
          print(paste("Current Task = ", CurrentTask))
          readline(prompt="Press [enter] to proceed")
      }
          
          AvgJobs=AvgJobs+dt*length(WaitingList);
    
      }
      
      ResponseTime = mean(Completion-Arrival); #average response time
      AvgJobs=AvgJobs/(tail(Completion,n=1)-Arrival[1]);
    
      R_time[scheduling_discipline,cnt]=ResponseTime;
      
    } # loop scheduling discipline
    
    } # loop lambda
    
    print(R_time)
    
    #plot(c(1:length(lambdas)), R_time["FCFS",])
    matplot(t(R_time), type = "l", xlab="Load (lambda*E[S])", ylab="Avg Response Time");
    legend("topleft", legend = rownames(R_time), col=1:4, pch=1) # optional legend
    axis(side=1,at=1:ncol(R_time),labels=lambdas)

} # end function

```


First, let us consider exponentially distributed service times. We obtain:

```{r eval=TRUE, include=TRUE}
knitr::opts_chunk$set(echo = FALSE)

set.seed(11); # Set seed for reproducibility
N = 1e3; # number of jobs to simulate;
mu=1;
Service= rexp(N,rate=mu); # Service times
plot_response_times(Service, 1/mu);

```

Then, let us assume that service times follows a Pareto$(x_m,\alpha)$ distribution where $x_m$ and $\alpha$ are the scale and shape parameters, respectively. Let us assume $\alpha=3$ and $x_m=1$. We obtain:

```{r eval=TRUE, include=TRUE}
knitr::opts_chunk$set(echo = FALSE)

xm=1;
alpha=3;
set.seed(11); # Set seed for reproducibility
N = 1e3; # number of jobs to simulate;
Service=xm/(runif(N)^(1/alpha));

plot_response_times(Service, xm*alpha/(alpha-1));

```
